---
title: "Calling functions in data.table"
author: "Alexis Sciau"
date: "22 avril 2020"
output:
  html_document:
    code_folding: hide
    df_print: kable
    theme: default
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading libraries, echo=TRUE, message=FALSE, warning=FALSE}
library(data.table)
library(lubridate)
library(ggplot2)
library(pracma)
library(fasttime)
library(xts)
library(microbenchmark)
theme_set(theme_minimal())
```

This file tests calling functions inside data.table objects. Reference to [Stack Overflow question](https://stackoverflow.com/questions/61344686/calling-user-defined-function-from-data-table-object/61345840?noredirect=1#comment108539582_61345840)

## Functions without if and/or loop statements

## Functions with loop statement
Functions with loop statements can't be called in a straightforward way in the brackets of the data.table ocject. Let's illustrate this fact with an example.

```{r example_bug_loop_1}
test <-data.table(a=c(1,2,3),b=c(4,5,6))
f_1 <- function(a,b){
  out <-0
  for (i in seq(1,a,1)){
    out <- out + b^(i) 
  }
  return(out)
}

expected_result<-c(4, 30, 258)

test[,f_1(a,b)]
```

```{r example_bug_loop_2}
f_2 <- function(a,b){
  out <-0
  for (i in 1:a){
    out <- out + b^(i) 
  }
  return(out)
}

test[,f_2(a,b)]
```

We understand from this simple tests that the function called in the data.table take as a parameter all the column a as a parameter, and not only the a on its row as expected. To solve this problem, to solutions are proposed. 

```{r solution_loop}
test[,c:=mapply(f_1, test[,a],test[,b])]

test[,d:=f(a,b),1L:nrow(test)]
```

We know test their speed !
```{r speed_test_loop}
a<-1:500
b<-500:1

test_1 <- data.table(a,b)
test_2 <- data.table(a,b)

bench <- microbenchmark(v_1 = test_1[,c:=mapply(f_1,test_1[,a],test_1[,b])],
                        v_2 = test_2[,c:=f_1(a,b),1L:nrow(test_2)],times=100L)
autoplot(bench)
bench
```

Both solutions are equivalent!