---
title: "5. Fitting Yield Curves"
author: "Alexis Sciau"
date: "08/04/2020"
output:
  html_document:
    code_folding: hide
    df_print: kable
    theme: default
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading libraries, echo=TRUE, message=FALSE, warning=FALSE}
library(data.table)
library(lubridate)
library(ggplot2)
library(pracma)
library(fasttime)
library(xts)
theme_set(theme_minimal())
```

## 5.1 Getting started
Though we do not have access to the same data as the book, ie a collection of US Treasury bond yields as at 29 May 2009, we retrieve from the [US treasury website](https://www.treasury.gov/resource-center/data-chart-center/interest-rates/Pages/TextView.aspx?data=yield "Daily Treasury Yield Curve Rates  from US Treasury") the daily treasury yield curve rates. Plotting this curve, we expect to get something close to the fig 5.2.

```{r US_yield_curve_example, message=FALSE, warning=FALSE}
yc_dt <- fread("../data/data_YC_US")
yc_dt[,index:=fastPOSIXct(index)]
yc_dt[,index:=as.Date(index)]
yc_xts <- as.xts(yc_dt)

tenors <- c(1/12,2/12,3/12,6/12,1,2,3,5,7,10,20,30)
colnames(yc_xts)<-tenors

rates <- yc_xts["20090528/20090603"]

daily_data <- as.data.table(melt(as.data.table(rates), measure.vars = colnames(rates)))
daily_data[,index_f := as.factor(index)]
daily_data[,tenors := as.numeric(as.character(variable))]
ggplot(daily_data,aes(x=tenors,y=value,color=index_f,group=index_f))+ scale_x_continuous(name="Tenors", limits=c(0, 30)) + geom_point() + geom_smooth(se=FALSE)
```

## 5.2 Yield Curves 101

### 5.2.1 Pure-Discount Bond Prices
We note that, in this paragraph, the author take the assumption that the interest rate can't be less than zero :

> "Some reflection reveals that the price should not exceed $1. Paying more than the future value essentially implies a negative interest rate [...]"

This fairly logical assumption does not hold today!

### 5.2.2 Spot Rates
We will illustrate the correspondance between pure-discount bond prices and spot rates. We start this exercise choosing arbitrary prices for Pure-Discount bonds.
```{r Discount prices and spot rates}
long_names <- c("Maturity","Time to maturity","Pure-Discount Bond Price","Return","Rate","Simply compounded spot rates","Continuously compounded spot rates")
short_names <- c("mat","time_to_mat","disc_f","return","rate","spot_sc","spot_cc")
dt_rates <- as.data.table(as.Date(c("2020-04-30","2020-06-30","2020-12-31","2021-12-31","2025-12-31")))
dt_rates[,time_to_mat:=V1-as.Date("2020-04-17")]
dt_rates[,disc_f:=c(0.9995,0.997,0.985,0.95,0.82)]
dt_rates[,return := 1/disc_f]
dt_rates[,rate:=(return -1)*100]
dt_rates[,spot_sc:=(return^(365/as.numeric(time_to_mat))-1)*100]
dt_rates[,spot_cc:=(-log(disc_f)/(as.numeric(time_to_mat)/365))*100]
colnames(dt_rates)<-long_names
dt_rates
dt_rates_plot <- melt(dt_rates,id = 2,measure=c(6,7))
colnames(dt_rates_plot)<-c("time_to_mat","Rates","value")
ggplot(dt_rates_plot,aes(x=time_to_mat,y=value,color=Rates)) + geom_point() + geom_line() + scale_x_continuous(name="Time to maturity (days)") + scale_y_continuous(name="Rate (%)")
```

The above graph illustrates the impact of the time to maturity on the difference between simply compounded and continuously compounded rates.

### 5.2.3 Par Yields
Let's now compute the par-yields for the set of bonds used just before.
```{r par_yields}
colnames(dt_rates)<-short_names
dt_rates[,df_cumsum:=cumsum(disc_f)]
dt_rates[,par_yield:=(1-disc_f)/df_cumsum*100]
```

### 5.2.5 Bringing it all together
Let's suppose the rates given by the US treasury are spot rates, ie zero-coupon bond rates. We will therefor try to plot the subsequent par-yield and implied-forward artes curves. Also, to ease-up the work, we will not consider tenors under 1 year. Finally, we do simple linear interpolation to complete the zero-bond curve every year.
```{r all_together, warning=FALSE}
all_rates <- as.data.table(yc_xts["20090529"])
all_rates <- melt(all_rates,id=1,measure=6:13)
all_rates[,tenors:=as.numeric(as.character(variable))]
all_rates[,spot_rates:=value]
all_rates[,c("index"):=NULL][,c("variable"):=NULL][,c("value"):=NULL]
all_tenors <- data.table(tenors=1:30)
all_rates <- merge(all_rates,all_tenors,all=TRUE)
all_rates <- as.data.table(na.approx(all_rates))
all_rates[,disc_f:=1/(1+spot_rates/100)^tenors]
all_rates[,df_cumsum := cumsum(disc_f)]
all_rates[,par_yield:=((1-disc_f)/df_cumsum)*100]
all_rates[,disc_f_30:=all_rates[tenors==30,disc_f]]
all_rates[,T_minus_s:=30-tenors]
all_rates[,forward_rates_mat_30:=((disc_f/disc_f_30)^(1/T_minus_s)-1)*100]
all_rates[,df_shift:=shift(disc_f,1,type="lead")]
all_rates[,forward_rates_1y:=(disc_f/df_shift-1)*100]
all_rates_plot<-melt(all_rates,id="tenors",measure=c("spot_rates","par_yield","forward_rates_mat_30","forward_rates_1y"))
ggplot(all_rates_plot,aes(x=tenors,y=value,color=variable)) + geom_point() + geom_line()
ggplot(all_rates,aes(x=tenors,y=disc_f))+ geom_point() + geom_line()
```


## 5.3 Curve Fitting

### 5.3.1 The Classic Approach
In this approach, several functions are taken as example to represent the four core consitutuents of the yield curve. We will go through the examples chosen in the book.s

#### Discount factors as negative exponential
The first example is to choose the negative exponential to represent the discount factors.

```{r negative_exp_df}
disc_f <- function(t,a=0.03,x=0){
  exp(-a*(t-x))
}
ex1_dt <- data.table(t=seq(0,30,0.1))
ex1_dt[,disc_factors_1 := disc_f(t,0.01)]
ex1_dt[,disc_factors_2 := disc_f(t,0.02)]
ex1_dt[,disc_factors_3 := disc_f(t,0.03)]
ex1_dt[,disc_factors_4 := disc_f(t,0.04)]
ex1_dt[,disc_factors_5 := disc_f(t,0.05)]
ex1_dt[,disc_factors_6 := disc_f(t,0.06)]
ex1_dt[,disc_factors_7 := disc_f(t,0.07)]
ex1_dt[,zero_coupon_1 := -log(disc_factors_1)/t*100]
ex1_dt[,zero_coupon_2 := -log(disc_factors_2)/t*100]
ex1_dt[,zero_coupon_3 := -log(disc_factors_3)/t*100]
ex1_dt[,zero_coupon_4 := -log(disc_factors_4)/t*100]
ex1_dt[,zero_coupon_5 := -log(disc_factors_5)/t*100]
ex1_dt[,zero_coupon_6 := -log(disc_factors_6)/t*100]
ex1_dt[,zero_coupon_7 := -log(disc_factors_7)/t*100]
ex1_dt_df_plot <- melt(ex1_dt,id="t",measure = c(2:8))
ex1_dt_zc_plot <- melt(ex1_dt,id="t",measure = c(9:15))
ggplot(ex1_dt_df_plot,aes(x=t,y=value,color=variable)) + geom_line()
ggplot(ex1_dt_zc_plot,aes(x=t,y=value,color=variable)) + geom_line() + scale_y_continuous(limits = c(0,10))
```

Numerically, we found back what is outlined in the book : using a negative exponential for representing discount factors leads to a flat zero coupon curve, equals to the parameter used in the negative exponential.

