---
title: "6. Modelling Yield Curves"
author: "Alexis Kenshuri"
date: "24 avril 2020"
output:
  html_document:
    code_folding: hide
    df_print: kable
    theme: default
    toc: yes
    toc_float: yes
    toc_depth: 4 
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r loading libraries, echo=TRUE, message=FALSE, warning=FALSE}
library(data.table)
library(lubridate)
library(ggplot2)
library(pracma)
library(fasttime)
library(xts)
library(YieldCurve)
library(plotly)
theme_set(theme_minimal())
```

## 6.1 Why a dynamic Yield-Curve Model ?

This chapter starts by some general observations about yield curves complexity.

```{r YC_dynamic_illustrations}
yc_dt <- fread("../data/data_YC_US")
yc_dt[,index:=fastPOSIXct(index)]
yc_dt[,index:=as.Date(index)]

tenors <- c(1/12,2/12,3/12,6/12,1,2,3,5,7,10,20,30)

colnames(yc_dt)<-c("index",tenors)

yc_dt_fig_61 <- melt(yc_dt,id="index")
yc_dt_fig_61[,tenor:=as.numeric(as.character(variable))]

yc_dt_fig_61[,index_f:=as.factor(index)]
ggplot(yc_dt_fig_61,aes(x=tenor,y=value,color=index_f))+geom_line(show.legend = FALSE)

yc_xts <- as.xts(yc_dt)
plot(yc_xts["1997/2011"])

summary(yc_dt)
yc_dt_fig_63 <- yc_dt_fig_61[tenor %in% c(0.25,0.5,1,2,3,5,7,10,20,30),mean_yield:=mean(value,na.rm=TRUE),by=tenor]
ggplot(yc_dt_fig_63,aes(x=tenor,y=mean_yield))+geom_line()+scale_y_continuous(limits = c(0,8))

yc_dt_fig_64 <- yc_dt_fig_61[tenor %in% c(0.25,0.5,1,2,3,5,7,10,20,30),mean_var:=var(value,na.rm=TRUE),by=tenor]
ggplot(yc_dt_fig_64,aes(x=tenor,y=mean_var))+geom_line()

yc_dt_col_txt <- fread("../data/data_YC_US")
yc_dt_col_txt[,index:=fastPOSIXct(index)]
yc_dt_col_txt[,index:=as.Date(index)]
cor_mat <- as.data.table(cor(yc_dt_col_txt[is.na(BC_2MONTH)==FALSE,-c(1:3,13)]))

cor_mat_plot<- cor_mat[,tenor_y:=colnames(cor_mat)]
cor_mat_plot <- melt(cor_mat_plot,id="tenor_y")
cor_mat[,tenor_y:=NULL]

cor_mat_plot[,tenor_yy:=factor(tenor_y,levels = rev(colnames(cor_mat)))]

ggplot(cor_mat_plot,aes(x=variable,y=tenor_yy,color=value, size=value)) + geom_point()

ggplot(cor_mat_plot,aes(x=variable,y=tenor_yy)) + geom_tile(aes(fill = value),colour="white") + scale_fill_gradient(low = "white", high = "steelblue")

yc_var_dt <- yc_dt_fig_61[,val_m_1:=shift(value,type="lag"),by=tenor]
yc_var_dt[,pct_change:=((value-val_m_1)/val_m_1)*100]
yc_var_dt <- yc_var_dt[,.(index,tenor,pct_change)]
yc_var_dt<-dcast(yc_var_dt,index~tenor,value.var = "pct_change")
colnames(yc_var_dt)<-colnames(yc_dt_col_txt)
cor_mat_change <- as.data.table(cor(yc_var_dt[is.na(BC_2MONTH)==FALSE & is.na(BC_2MONTH)==FALSE,-c(1:4,13)]))

cor_mat_plot<- cor_mat_change[,tenor_y:=colnames(cor_mat_change)]
cor_mat_plot <- melt(cor_mat_plot,id="tenor_y")
cor_mat[,tenor_y:=NULL]

cor_mat_plot[,tenor_yy:=factor(tenor_y,levels = rev(colnames(cor_mat)))]

ggplot(cor_mat_plot,aes(x=variable,y=tenor_yy,color=value, size=value)) + geom_point()

ggplot(cor_mat_plot,aes(x=variable,y=tenor_yy)) + geom_tile(aes(fill = value),colour="white") + scale_fill_gradient(low = "white", high = "steelblue")
```

Also, it states that **discount factors must be decreasing to avoid negative-rates and thus arbitrage opportunities**: this is not obvious/true anymore!

### 6.1.3 Risk Premia

We will try to understand better risk premia by recreating figure 6.7. The process to achive this is the following :
* get the rate and subset the same as in the book
* get NS parameters
* construct spot rates --> Nous faisons ici l'hypothèse que les données sont déjà des spot rates et pas des yiels pour séconomiser le besoin de les transformer
* construct discount factors
* construct forward rates
* compute risk premia

```{r risk_premia}
rp<-to.monthly(na.omit(yc_xts["1997/2011"][,-c(1,2,12)]),indexAt = "lastof", OHLC=FALSE)
matu<-as.numeric(colnames(rp))

NSParameters <- Nelson.Siegel(rate = rp, maturity = matu)
tenor <- seq(0.25,20,0.25)
rp_dt <- as.data.table(NSrates(NSParameters,tenor))
colnames(rp_dt) <- c("index",tenor)
rp_dt <- melt(rp_dt,id="index")
rp_dt[,tenor:=as.numeric(as.character(variable))]
rp_dt[,spot:=value]
rp_dt[,df:=1/(1+spot/100)^tenor]
rp_dt[,df_1:=.SD[tenor==1,df],by=index]
rp_dt[,df_p_1:=shift(df,type="lead",n=4),by=index]
rp_dt[,forward := ((df_1/df_p_1)^(1/tenor)-1)*100]
rp_dt[,future_spot:=shift(spot,n=12,type="lead")]
rp_dt[index>as.Date("2011-01-01"),future_spot:=NA]
rp_dt[,risk_premia:=(forward-future_spot)*100]
rp_dt[,index_f:=as.factor(index)]
rp_dt[,mean_rp:=mean(risk_premia,na.rm=TRUE),by=tenor]
ggplot(rp_dt,aes(x=tenor)) + geom_line(aes(y=risk_premia,color=index_f),show.legend = FALSE) + geom_line(aes(y=mean_rp),color="black")
```

We end up with a figure completely different from the fig 6.7 in the book... The conclusion, though, in the same : **forward rates are above realized futures rates**.

## 6.3 Statistical digression

Let's apply the PCA on the US rates data.

```{r pca}
pca <- prcomp(rp)
summary(pca)

pca_dt <- as.data.table(pca$x)
pca_dt <- pca_dt[,c(1:3)]
pca_dt[,t:=index(rp)]
pca_dt_plot <- melt(pca_dt,id="t")
ggplot(pca_dt_plot,aes(x=t,y=value, color=variable)) + geom_line() + ggtitle("Three Largest Principal Components") 

pca_loadings_dt <- as.data.table(pca$rotation)
pca_loadings_dt <- pca_loadings_dt[,c(1:3)]
pca_loadings_dt[,t:=c(0.25,0.5,1,2,3,5,7,10,20)]
pca_loadings_dt_plot<-melt(pca_loadings_dt,id="t")
ggplot(pca_loadings_dt_plot,aes(x=t,y=value, color=variable)) + geom_line()  + ggtitle("PCA loadings")
```

Although satisfactory, we will restrain from going further and invest some time on Mathematics theory.

### Digression 1 : Eigen Values et Eigen Vectors

See Annex A.4.3

Also :
* [Cours Supinfo](https://www.supinfo.com/cours/1LAL/chapitres/04-diagonalisation-matrices-carrees)
* [Cours univ Paris](https://www.math.univ-paris13.fr/~schwartz/L2/diag.pdf)


### Digression 2 : Proxies for PCA factors

In p173, the shaded box provides us with proxies for the level, slope and curvature. Let's try to compute correlation to see if these proxies are satisfying.

```{r PCA_proxies}
rp_dt <- as.data.table(rp)
colnames(rp_dt)<-c("t","M3","M6","Y1","Y2","Y3","Y5","Y7","Y10","Y20")
rp_dt[,level:=Y20]
rp_dt[,slope:=Y20-M3]
rp_dt[,curvature:=2*Y5 - (M3+Y20)]
cor(pca_dt[,PC1],rp_dt[,level])
cor(pca_dt[,PC2],rp_dt[,slope])
cor(pca_dt[,PC3],rp_dt[,curvature])
```

The correlation between proxies and PCA factors is good but not as good as the one found in the book.

## 6.4 Model Examples

### 6.4.1 Toy Model

Let's tru to fit the toy model to our data.

```{r toy_model}
PC1_model <- ar.ols(pca_dt[,PC1],order.max=1,intercept = TRUE,demean = FALSE)
PC2_model <- ar.ols(pca_dt[,PC2],order.max=1,intercept = TRUE,demean = FALSE)
PC3_model <- ar.ols(pca_dt[,PC3],order.max=1,intercept = TRUE,demean = FALSE)

map_dt <- as.data.table(rp)
colnames(map_dt)<-c("t","M3","M6","Y1","Y2","Y3","Y5","Y7","Y10","Y20")
map_dt <- melt(map_dt,id="t")
colnames(map_dt) <- c("t","tenor","yield")
map_dt[,yield_n:=shift(yield,1,type="lead"),by=tenor]
map_dt[,yield_diff:=yield_n-yield]
map_dt<-merge(map_dt,pca_dt)

mapping <- lm(yield_diff~PC1+PC2+PC3-1,map_dt)

parameters<- data.table(map_beta=mapping$coefficients,dyn_alpha=c(PC1_model$x.intercept,PC2_model$x.intercept,PC3_model$x.intercept),dyn_gamma=c(PC1_model$ar,PC2_model$ar,PC3_model$ar))
print(parameters)
```

The parameters that we estimated for the model are far from the ones given in the book. We will nevertheless try to estimate the in-sample fit of our model.

```{r toy_model_in_sample_fit}
avg_yield_by_tenor <- map_dt[,avg_yield:=mean(yield),by=tenor][1:9,.(tenor,avg_yield)]
map_dt[,estimated_yield_aux:=yield+parameters[1,map_beta]*PC1+parameters[2,map_beta]*PC2+parameters[3,map_beta]*PC3]
map_dt[,estimated_yield := shift(estimated_yield_aux,1,type="lag"),by=tenor]
avg_est_yield_by_tenor <- map_dt[,avg_est_yield:=mean(estimated_yield,na.rm=TRUE),by=tenor][1:9,.(tenor,avg_est_yield)]
toy_plot <- data.table(t=c(0.25,0.5,1,2,3,5,7,10,20),avg_yield_by_tenor[,avg_yield],avg_est_yield_by_tenor[,avg_est_yield])
colnames(toy_plot)<-c("t","Actual_Average_Yield","Estimated_Average_Yield")
toy_plot<-melt(toy_plot,id="t")
ggplot(toy_plot,aes(x=t,y=value,color=variable))+geom_line()+scale_y_continuous(limits = c(1,6))
```

In the above, we only tested the fit of the mapping (hypothesis A3), it thus seems logical that this fit is almost perfect (on average). We didn't test for the fit of the dynamic model. I understand that this is the same that it was done in the book p176, thus making the difference observed hardly explainable.

```{r toy_model_est_errors}
map_dt[,est_error:=estimated_yield-yield]
est_error_plot <- as.data.table(map_dt[,.(t,tenor,est_error)])
est_error_plot<-est_error_plot[-c(1:9)]
est_error_plot[,t_f:=as.factor(t)]
est_error_plot[,tenor_num:=c(0.25,0.5,1,2,3,5,7,10,20)]
ggplot(est_error_plot,aes(x=tenor_num))+geom_line(aes(y = est_error,color=t_f),show.legend = FALSE)
plot_ly(est_error_plot, x = ~tenor_num, y = ~t_f, z = ~est_error, type = "scatter3d", mode = "lines")
```

### 6.4.2 A Complex Example